\noindent

Firstly, I can't thank Dr. Tushar Sharma enough. He's been an incredible supervisor since before I even joined Dalhousie. The trust he placed in me was amazing! Beyond research guidance, he helped with teaching opportunities and external projects. What makes him special is how approachable he is - whether it's academic or personal issues, his door was always open. While I wasn't always a perfect student and made mistakes along the way, he helped me learn and grow from each experience. His endless encouragement, thoughtful guidance, and understanding nature made this thesis possible. Tushar, you're the best mentor I could've asked for - thank you for everything!

To the members of our Software Maintenance and Analytics Research Team (SMART) lab - Mootez Saad, Saurabhsingh Rajput and Gautam Shetty, I consider myself to be extremely fortunate to have met you all and got the opportunity to work alongside such great minds. I am also grateful for all the interactive lab meetings we had, from brainstorming research ideas to dreaming up (hilariously failed) startup plans. Those late night discussions about everything under the sun - football, politics, food recipes, and those endless jokes, are memories I'll always cherish. I would like to specially thank my exceptionally talented and hard working friend and peer, Mootez, for time and again helping me with my research providing his insightful feedback and inputs. These suggestions had a significant impact in polishing my work. Working with him was truly inspirational. I would also like to thank the previous members and interns of our lab. It is an honor to become an alumni of this lab.

To Saurabh and Gautam, this section doesn't have sufficient space to accommodate everything that they have done for me. We three have become brothers since day one. Every memory in Halifax involves these two - from last minute assignment rushes and technical discussions to exploring the city together, watching football matches, and impromptu trips. Saurabh's laid-back attitude and reassuring presence with deep technical knowledge helped me navigate through challenging times. Gautam's wisdom and thoughtful insights, both on and off work discussions have been invaluable. I am also indebted to Gautam for his valuable contribution as an author in my first research paper.

I am deeply thankful to all the members of the Intelligent Automation in Software EngineeRing (RAISE) lab - Mehil B Shah, Asif Samir, Usmi Mukherjee, Sigma Jahan and Riasat Mahbub. I would like to acknowledge and appreciate my very good friends Usmi, Mehil and Sigma for supporting me and cheering me both professionally and personally. Mehil is very vocal and I admire his innocence and authentic presence. Sigma is a multi-tasker and I really think she has invisible octopus hands. Finally, I would like to thank Usmi for being there as a friend and a strong support system throughout the journey. She is one of the best people I ever met. Also, she is the only friend from my hometown, Kolkata. Celebrating Durga Puja as \textit{probashi bangalis} (expatriate bengalis) will always be a great experience.

I would like to extend my sincere gratitude towards Dalhousie University and the Faculty of Computer Science for providing me with this splendid learning opportunity. I would like to thank Dr. Michael McAllister for promptly resolving a major administrative issue I was facing. I would also like to thank the department for awarding me with graduate bursaries twice. I am very fortunate to have received funding support from \texttt{NRC IRAP}, \texttt{Lab2Market} and \texttt{Mitacs Accelerate}. The internships and projects enabled by this funding provided valuable opportunities to apply my research in practical, real-world settings. The defense

This graduate journey would not have been possible without the constant support and encouragement from my incredible parents, brother and other close family members, who have stood by me every step of the way. I am deeply grateful for their belief in me and the strength they have given me throughout this journey. I would also like to take a moment to thank all my friends back home, especially Jessica Banik and Swagnik Banerjee, whose friendship and support have been invaluable during this transformative phase of my life. 

Lastly, I sincerely thank everyone who supported me throughout this research journey. Their encouragement, advice, and assistance have been invaluable, and I am truly grateful.